---
title: "Reproducible Research - Assignment 1"
author: "Arun Panangatt"
date: "14 May 2015"
output: html_document
---
## Overarching settings
```{r}
echo = TRUE ### to make all code visible
library(ggplot2)

```
## Loading data and pre-processing
```{r}
data <- read.table('activity.csv',sep = ",",header = TRUE, na.strings ="NA",
                   colClasses = c('integer','Date','factor'))
```
### To clear rows with NA
```{r}
new.data <- na.omit(data)
```
###  Plot of total number of steps taken per day
```{r}
total.steps <- tapply(new.data$steps, new.data$date, FUN = sum)
plot1 <- ggplot(new.data, aes(date, steps)) + geom_bar(stat = "identity",fill='red',binwidth = .5) +
        labs(title = " Total Number of Steps Taken Each Day",x = "Date", y = " Number of Steps")
print(plot1)
```


### Mean and Median of the total numberof steps taken per day
```{r}
summary(total.steps)
```

### Average daily actitivy pattern plot
```{r}
averages <- aggregate(new.data$steps, list(interval = as.numeric(as.character(new.data$interval))), FUN = "mean")
names(averages)[2] <- "Avg.Steps"

plot2 <- ggplot(averages, aes(interval, Avg.Steps)) + geom_line(color = "black", size = 0.7) + labs(title = "Time Series Plot : 5-minute Intervals", x = "5-minute intervals", y = "Average Number of Steps Taken")
print(plot2)
```

### 5 minute interval with the highest average number of steps
```{r}
averages[averages$Avg.Steps == max(averages$Avg.Steps),]
```

### Interval named 835 with 206 steps at index position 104 is the highest.

### Imputing missing values

### Total number of incomplete cases

```{r}
sum(!complete.cases(data))
```


### Imputing missing values with mean for 5 minute interval

```{r}
impData <- data 
for (i in 1:nrow(impData)) {
    if (is.na(impData$steps[i])) {
        impData$steps[i] <- averages[which(impData$interval[i] == averages$interval), ]$Avg.Steps
    }
}
```

### Verification of missing values having been filled in

```{r}
sum(!complete.cases(impData))
```


### Histogram with imputed data
```{r}
plot3 <- ggplot(impData, aes(date, steps)) + geom_bar(stat = "identity", fill="red",binwidth = .5) +
    labs(title = "Total Number of Steps Taken Each Day (Imputed Data)",x = "Date", y = " Number of Steps")
print(plot3)
```
### Calculation of mean and median of imputed data
```{r}
totalsteps.impute <- tapply(impData$steps, impData$date, FUN = sum)
summary(totalsteps.impute)
```

### It can be noted that the Median has increased in the imputed data , but the Mean remanins the same

### Differences in activity patterns between Weekdays and Weekends

### Creation of new variable indicating whether a given date is Weekday or Weekend
```{r}
impData$weekdays <- factor(format(impData$date, "%A"))
levels(impData$weekdays)
```
```{r}
levels(impData$weekdays) <- list(weekday = c("Monday", "Tuesday",
                                             "Wednesday", 
                                             "Thursday", "Friday"),
                                 weekend = c("Saturday", "Sunday"))
levels(impData$weekdays)
```
### Graph containing  time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r}
new.averages <- aggregate(impData$steps, 
                      list(interval = as.numeric(as.character(impData$interval)), 
                           weekdays = impData$weekdays),
                      FUN = "mean")
names(new.averages)[3] <- "meanOfSteps"
library(lattice)
plot4 <- xyplot(new.averages$meanOfSteps ~ new.averages$interval | new.averages$weekdays, 
       layout = c(1, 2), type = "l", 
       xlab = "Interval", ylab = "Number of steps")
print(plot4)
```
### Difference between Weekdays and Weekends
```{r}
table(impData$weekdays)
```



